---
title: "draft-2"
author: "Cameron Pellett"
date: "02/12/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.height = 9, fig.width = 9)
```

```{r packages}
library(tidyverse)
library(patchwork)
library(broom)
library(ggbeeswarm)

theme_set(theme_grey()+theme(panel.background = element_rect(fill = "grey96")))
```

```{r functions}

#summarises mean, sd, se, ci for bar charts
mean_ci_summary <- function(data, groups, variable, log10 = FALSE){
  
  data <- data|>ungroup()
  
  for (i in 1:length(groups)) {
    
    data <- data|>
      group_by(.data[[ groups[i] ]], .add = TRUE)
    
  }
  
  if(log10 == TRUE){
    
    data|>
    mutate(varna = case_when(is.na(.data[[variable]]) | is.nan(.data[[variable]]) | is.infinite(.data[[variable]]) ~ as.integer(1),
                             TRUE ~ NA_integer_))|>
      summarise(mean = mean(log10(.data[[variable]]), na.rm = TRUE),
                sd = sd(log10(.data[[variable]]), na.rm = TRUE),
                n = n() - sum(varna, na.rm = TRUE))|>
      mutate(se = sd / sqrt(n),
             ci = se * qt(1 - 0.05 / 2, n-1))
    
  }else {
    
    data|>
      mutate(varna = case_when(is.na(.data[[variable]]) | is.nan(.data[[variable]]) | is.infinite(.data[[variable]]) ~ as.integer(1),
                             TRUE ~ NA_integer_))|>
      summarise(mean = mean(.data[[variable]], na.rm = TRUE),
                sd = sd(.data[[variable]], na.rm = TRUE),
                n = n()  - sum(varna, na.rm = TRUE))|>
      mutate(se = sd / sqrt(n),
             ci = se * qt(1 - 0.05 / 2, n-1))
    
  }
}

pvalue_fun <- function(p.value){
  if(is.na(p.value)){
    NA_character_
  }else if(p.value > 0.05){
    "> 0.05"
  }else if(p.value < 0.05 & p.value > 0.01){
    "< 0.05"
  }else if(p.value < 0.01 & p.value > 0.001){
    "< 0.01"
  }else if(p.value < 0.001){
    "< 0.001"
  }else {
    "something went wrong"
  }
}

pvalue_star <- function(p.value){
  if(is.na(p.value)){
    NA_character_
  }else if(p.value > 0.05){
    "."
  }else if(p.value < 0.05 & p.value > 0.01){
    "*"
  }else if(p.value < 0.01 & p.value > 0.001){
    "**"
  }else if(p.value < 0.001){
    "***"
  }else {
    "something went wrong"
  }
}

stat_paste_fun <- function(stat, log = FALSE){
  if(log != TRUE){
  paste0(str_remove_all(stat[["method"]][[1]], "\n\t"), ": ", 
         str_remove_all(names(stat[["statistic"]]), "Kruskal-Wallis "), " = ",
         round(stat[["statistic"]][[1]],1), ", ",
         if(length(stat[["parameter"]]) == 0)
         {""
         }else if(is.na(stat[["parameter"]])){
           ""
         }else if(length(stat[["parameter"]]) > 0){
           paste0("df = ", round(stat[["parameter"]], 0), ", ")
         } ,
         "p-value ",  pvalue_fun(stat[["p.value"]]))
  }else{
    paste0(str_remove_all(stat[["method"]][[1]], "\n\t"), " (log y): ", 
           str_remove_all(names(stat[["statistic"]]), "Kruskal-Wallis "), " = ",
           round(stat[["statistic"]][[1]],1), ", ",
           if(length(stat[["parameter"]]) == 0)
           {""
           }else if(is.na(stat[["parameter"]])){
             ""
           }else if(length(stat[["parameter"]]) > 0){
             paste0("df = ", round(stat[["parameter"]], 0), ", ")
           } ,
           "p-value ",  pvalue_fun(stat[["p.value"]]))
  }
}

paste_anova_fun <- function(anova, log = FALSE){
  
  if(log != TRUE){
  paste0("ANOVA: ", "F value = ", round((anova$`F value`)[1], 2), ", ", "p-value ", pvalue_fun(anova$`Pr(>F)`[1]))
  }else {
    paste0("ANOVA (log y): ", "F value = ", round((anova$`F value`)[1], 2), ", ", "p-value ", pvalue_fun(anova$`Pr(>F)`[1])) 
  }
}

lmR2adj <- function(lm){
  
  R2 <- 1-((sum(residuals(lm)^2))/
             (sum((lm$model[[1]] - mean(lm$model[[1]]))^2)))
  
  1- ((1 - R2) * (length(lm$model[[1]] ) - 1)/
                 (length(lm$model[[1]] ) - length(lm$coefficients) - 1))
}

lmR2 <- function(lm){
  
  1-((sum(residuals(lm)^2))/
             (sum((lm$model[[1]] - mean(lm$model[[1]]))^2)))
}

glmR2 <- function(glm, adj = FALSE){
  
  residuals <- glm[["y"]] - glm[["fitted.values"]]
  
  R2 <- 1-((sum(residuals^2))/
             (sum((glm$model[[1]] - mean(glm$model[[1]]))^2)))
  
  if(adj != TRUE){
    
  return(R2)
    
  }else {
    
    1- ((1 - R2) * (length(glm$model[[1]] ) - 1)/
                 (length(glm$model[[1]] ) - length(glm$coefficients) - 1))
  }
}



pairwise_diff_comps <- function(pairwise_pv_matrix){
  
  pairwise_diff <- as_tibble(pairwise_pv_matrix, rownames = "vars")|>
  pivot_longer(2:last_col(), names_to = "var.y", values_to = "p.value")|>
  na.omit()|>
  filter(p.value < 0.05)|>
  select(1,2)|>
  unite("comp", c(1,2), sep = " ~ ")
  
  comp_diffs <- ""
  
  for (i in 1:nrow(pairwise_diff)) {
    
    if(i == nrow(pairwise_diff)){
      comp_diffs <- paste0(c(comp_diffs, pairwise_diff[[i,1]]), collapse = ", and ")
    }else if(i != 1){
      comp_diffs <- paste0(c(comp_diffs, pairwise_diff[[i,1]]), collapse = ", ")
    }else{
      comp_diffs <- pairwise_diff[[i,1]]
    }
    
  }
  
  return(comp_diffs)
}

plot_assumptions <- function(model, obs, bins = 30){
        #create tibble for use in ggplot of model assumptions
        df <- tibble(
                observations = obs,
                fitted = fitted(model),
                residuals = resid(model),
                standardised_residuals = sqrt(resid(model)^2),
                RMSresiduals = sqrt(mean(resid(model)^2)))
        
        
        a <- ggplot(df, aes(fitted, residuals))+
                geom_point()+
                geom_abline(aes(slope = 0, intercept = 0))
                #labs(title = "Heteroscedasticity", subtitle = "Check trends not captured by model")
        
        b <- ggplot(df, aes(residuals))+
                geom_histogram(bins = bins)
        
        c <- ggplot(df, aes(sample = residuals))+
                geom_qq()+
                geom_qq_line()#+
                #labs(title = "Quantile-quantile plot", subtitle = "check normality of residuals")
        
        d <- ggplot(df, aes(observations, fitted))+
                geom_point()+
                geom_abline()#+
                #labs(subtitle = "Check accuracy of model across whole dataset")
        
        plot <- a+b+c+d+plot_layout(ncol = 2, nrow = 2)+plot_annotation(tag_levels = "a")
        return(plot)
} 


pairwise_plot_layer <- function(pairwise.test, ymin, ymax, xmin, xmax, x, y, scale, data){

pairwise <- pairwise.test[["p.value"]]|>
  as_tibble(rownames = "rownames")|>
  rowwise()|>
  mutate(across(2:last_col(), pvalue_star),
         across(2:last_col(), ~replace_na(.x, "")))

title = tibble(ymin = ymin, ymax = ymax, xmin = xmin, xmax = xmax, 
               x = x, y = y, scale = scale,
               ncol = (ncol(pairwise)-1)/2,
               labels = paste0("Pairwise ",
                              pairwise.test[["method"]],
                              " (", pairwise.test[["p.adjust.method"]],
                              " adj.)"))|>
  mutate(labely = y + ((ymax-ymin)* 1/(16/scale)),
         labelx = x + ((xmax-xmin)* ncol/(16/scale)))
  

ylabel <- tibble(ymin = ymin, ymax = ymax, xmin = xmin, xmax = xmax, 
       x = x, y = y, scale = scale,
       nrow = 1:nrow(pairwise),
       labels = pull(pairwise, rownames))|>
  mutate(labely = y - ((ymax-ymin)* nrow/(16/scale)),
         labelx = x)

xlabel <- tibble(ymin = ymin, ymax = ymax, xmin = xmin, xmax = xmax, 
                 x = x, y = y, scale = scale,
       ncol = 1:(ncol(pairwise)-1),
       labels = colnames(pairwise)[-1])|>
  mutate(labely = y,
         labelx = x + ((xmax-xmin)* ncol/(16/scale)))


xypvalue <- tibble(ymin = ymin, ymax = ymax, xmin = xmin, xmax = xmax, 
                   x = x, y = y, scale = scale,
       nrow = rep(1:nrow(pairwise), each = ncol(pairwise)-1),
       ncol = rep(1:(ncol(pairwise)-1), nrow(pairwise)),
       labels = NA_character_)|>
  mutate(labely = y- ((ymax-ymin)* nrow/(16/scale)),
         labelx = x + ((xmax-xmin)* ncol/(16/scale)))


for (i in 1:nrow(xypvalue)) {

 nrow <-  (xypvalue[i,])|>
    pull(nrow)
 ncol <-  (xypvalue[i,])|>
   pull(ncol)
 
 xypvalue[[i, 10]] <- pairwise[[nrow, (ncol+1)]]
}

pairwise_label <- ylabel|>bind_rows(xlabel, xypvalue, title)

ex.pairwise_label <- pairwise_label|>
  bind_rows(tibble(
    ymin = rep(NA_real_, (nrow(data)-nrow(pairwise_label)))
    ))

return(ex.pairwise_label)
}
```


```{r data import and clean}

data <- read_csv("Spiking_data.csv")

virus_char <- read_csv("virus_characteristics.csv")|>
  pivot_longer(2:12, names_to = "virus", values_to = "value")|>
  pivot_wider(names_from = "X1", values_from = "value")|>
  mutate(virus = str_replace_all(virus, c("SARS" = "N1")))

clean <- data|>
  select(-Sample_code)|>
  rename("method" = Method, "spike" = Spike,
         "water_type" = Water_type,
         "volume" = Volume, "replicate" = Replicate)|>
  pivot_longer(cols = contains("rec"), names_to = "virus", values_to = "rec")|>
  mutate(virus = str_remove_all(virus, "_Rec%| Rec%"))|>
  mutate(rec = case_when((spike == "U" & virus %in% c("N1", "Phi6", "NoVGII") )~ NA_real_,
                         spike == "S" & water_type == "DW" & virus %in% c("AdV", "PMMoV", "CrAss") ~ NA_real_,
                         TRUE ~ rec))|>
  filter(!method %in% c("Pellet", "Pellet_BE"))|>
  left_join(virus_char, by = "virus")|>
  mutate(rec = case_when(round(rec,2) <= 0 ~ NA_real_,
                         TRUE ~ round(rec,2)),
         shape = factor(shape),
         water_type = factor(water_type),
         enveloped = factor(enveloped),
         size_nm = as.double(size_nm),
         method = factor(method),
         samples_spiked = str_replace_all(samples_spiked, c("yes" = "S",
                                                            "no" = "U")))|>
  na.omit()


pellet <- data|>
  select(-Sample_code)|>
  rename("method" = Method, "spike" = Spike,
         "water_type" = Water_type,
         "volume" = Volume, "replicate" = Replicate)|>
  pivot_longer(cols = contains("rec"), names_to = "virus", values_to = "rec")|>
  mutate(virus = str_remove_all(virus, "_Rec%| Rec%"))|>
  mutate(rec = case_when((spike == "U" & virus %in% c("N1", "Phi6", "NoVGII") )~ NA_real_,
                         spike == "S" & water_type == "DW" & virus %in% c("AdV", "PMMoV", "CrAss") ~ NA_real_,
                         TRUE ~ rec))|>
  filter(method %in% c("Pellet", "Pellet_BE", "PEG", "BE-PEG"))|>
  left_join(virus_char, by = "virus")|>
  mutate(rec = case_when(round(rec,2) <= 0 ~ NA_real_,
                         TRUE ~ round(rec,2)),
         shape = factor(shape),
         water_type = factor(water_type),
         enveloped = factor(enveloped),
         size_nm = as.double(size_nm),
         method = factor(method))|>
  na.omit()|>
  mutate(BE = case_when(str_detect(method, "BE") ~ "Beef extract",
                        TRUE ~ "No beef extract"))

```



```{r notes}

notes <- (tibble(estimate.x = "p-value: < 0.001 [***]; < 0.01 [**]; < 0.05 [*]; > 0.05 [.]"))
```


```{r full dataset plot, fig.height = 13, fig.width=9}
ww_all <- clean|>
  filter(water_type == "WW")|>
  mutate(volume = fct_reorder(.f = factor(paste0(volume, "ml WW")), .x = volume, .fun = median),
         shape1 = case_when(method == "Amicon" & volume == "20ml" ~ as.integer(32),
                            virus %in% c("AdV", "PMMoV", "CrAss") ~ as.integer(32),
                            TRUE ~ as.integer(16)),
         shape2 = case_when(method == "Amicon" & volume == "20ml" ~ as.integer(16),
                            TRUE ~ as.integer(32)),
         shape3 = case_when(virus %in% c("AdV", "PMMoV", "CrAss") & volume != "20ml" ~ as.integer(16),
                            TRUE ~ as.integer(32)))|>
  ggplot(aes(y = rec, x =method))+
  geom_beeswarm(aes(shape = shape1), priority = "ascending", cex = 4, show.legend = F)+
  geom_beeswarm(aes(shape = shape2), priority = "ascending", cex = 12, show.legend = F)+
  geom_beeswarm(aes(shape = shape3), priority = "ascending", cex = 6, show.legend = F)+
  geom_beeswarm(aes(shape = shape1), priority = "ascending", cex = 4, show.legend = F, size = 0.1, colour = "white")+
  geom_beeswarm(aes(shape = shape2), priority = "ascending", cex = 12, show.legend = F, size = 0.1, colour = "white")+
  geom_beeswarm(aes(shape = shape3), priority = "ascending", cex = 6, show.legend = F, size = 0.1, colour = "white")+
  scale_shape_identity()+
  scale_y_continuous(sec.axis = sec_axis( trans=~.*1, name="Virus", labels = NULL, breaks = NULL))+
  facet_grid(samples_spiked + virus~ volume, scale = "free", space = "free_x")+
  labs(x = "Concentration method", y = "Recovery (%)")

dw_all <- clean|>
  filter(water_type == "DW")|>
  mutate(volume = fct_reorder(.f = factor(paste0(volume, "ml DW")), .x = volume, .fun = median),
         shape1 = case_when(method == "Amicon" & volume == "20ml" ~ as.integer(32),
                            virus %in% c("AdV", "PMMoV", "CrAss") ~ as.integer(32),
                            TRUE ~ as.integer(16)),
         shape2 = case_when(method == "Amicon" & volume == "20ml" ~ as.integer(16),
                            TRUE ~ as.integer(32)),
         shape3 = case_when(virus %in% c("AdV", "PMMoV", "CrAss") & volume != "20ml" ~ as.integer(16),
                            TRUE ~ as.integer(32)))|>
  ggplot(aes(y = rec, x =method))+
  geom_beeswarm(aes(shape = shape1), priority = "ascending", cex = 4, show.legend = F)+
  geom_beeswarm(aes(shape = shape2), priority = "ascending", cex = 12, show.legend = F)+
  geom_beeswarm(aes(shape = shape3), priority = "ascending", cex = 6, show.legend = F)+
  geom_beeswarm(aes(shape = shape1), priority = "ascending", cex = 4, show.legend = F, size = 0.1, colour = "white")+
  geom_beeswarm(aes(shape = shape2), priority = "ascending", cex = 12, show.legend = F, size = 0.1, colour = "white")+
  geom_beeswarm(aes(shape = shape3), priority = "ascending", cex = 6, show.legend = F, size = 0.1, colour = "white")+
  scale_shape_identity()+
  scale_y_continuous(sec.axis = sec_axis( trans=~.*1, name="Virus", labels = NULL, breaks = NULL))+
  facet_grid(samples_spiked + virus~ volume, scale = "free", space = "free_x")+
  labs(x = "Concentration method", y = "Recovery (%)")

(ww_all + theme(axis.title.x = element_blank())) / dw_all + plot_layout(heights = c(1.1, 0.8))
```


Figure 1: recovery percentage for all 11 spiked (S) and unspiked (U) viruses grouped by wastewater (WW), de-ionised water (DW), sample volume (ml), and concentration method. Points have been offset on the horizontal axis to avoid over plotting.

<br>

Table1: generalised linear model with Gamma errors predicting recovery of spiked virus. Pellet methods were excluded from the model due to non-standard starting volumes. The model results include the variable coefficient on a log scale indicating its effect (positive numbers indicating increased recovery), followed by the standard error in brackets, and a significance code (`r notes`).


```{r overview multiple linear model}
glm <- glm(rec ~ volume + method + water_type + shape + enveloped + genome, data = clean, family = Gamma(link = "log"))

lm_table <- tidy(glm)|>
  mutate(term = str_replace_all(term, c("[(]Intercept[)]" = "Intercept (Amicon; DW; Head-tail; DNA)",
                                        "volume" = "Volume",
                                        "methodAS" = "Method: AS",
                                        "methodBE-PEG" = "Method: BE-PEG",
                                        "methodIP" = "Method: IP",
                                        "methodPEG" = "Method: PEG",
                                        "size_nm" = "Virus genome size (nm)",
                                        "water_typeWW" = "Water type: WW",
                                        "shapespherical" = "Virus shape: spherical",
                                        "shapehelical" = "Virus shape: helical",
                                        "shapeicosahedral" = "Virus shape: icosahedral",
                                        "envelopedyes" = "Enveloped: TRUE",
                                        "genomeRNA" = "Virus genome: RNA")),
         across(2:4, ~round(.x, 3)))|>
  rowwise()|>
  mutate(p.value = pvalue_star(p.value))|>
  unite(col = "estimate", c("estimate", "std.error"), sep = " (")|>
  unite(col = "estimate", c(2,4), sep = ") ")|>
  select(-statistic)


aicr2 <- tibble(
  term = c("AIC", "R-squared", "Adj. R-squared"), 
  estimate = c(AIC(glm), glmR2(glm), glmR2(glm, adj = TRUE))
)|>
  mutate(across(2, ~as.character(round(.x, 3))))



lm_table_fin <- lm_table|>
  bind_rows(tibble(term = ""))|>
  bind_rows(aicr2)|>
  mutate(across(1:2, ~replace_na(.x, "")))

knitr::kable(lm_table_fin, col.names = c("Variables", "Coefficients"))

```


<br>

```{r water type comp, fig.height = 6}
wt_comp <- clean|>
  filter(spike == "S" & samples_spiked == "S")

wt_comp_t.t <- tibble(label = stat_paste_fun(t.test(log(rec) ~ water_type, data = wt_comp), log = TRUE),
       row = 1)

(wt_comp|>
  mutate(row = row_number())|>
  left_join(wt_comp_t.t, by = "row")|>
  ggplot(aes(water_type, rec))+
  geom_boxplot(width = 2, outlier.alpha = 0)+
  geom_beeswarm(alpha = 0.5)+
  geom_text(aes(x = 1.5, y = max(rec) + (0.07 * (max(rec) - min(rec))),
                label = label), size = 3)+
  labs(x = "Water type",
       y = "Recovery (%)")) +
  
  (wt_comp|>
  ggplot(aes(water_type, rec))+
  geom_boxplot(width = 2, outlier.alpha = 0)+
  geom_beeswarm(alpha = 0.2, cex = 3)+
  facet_wrap(~virus, scale = "free")+
  labs(x = "Water type",
       y = "Recovery (%)") +
  theme(axis.title.y = element_blank())) +
  
  plot_layout(widths = c(1.3, 0.7)) +
    plot_annotation(tag_levels = "a")
  
```


Figure 2: De-ionised water has greater recovery of spiked virus compared to wastewater. Comparisons were made with a Welch Two Sample t-test after log transformation of recovery. Panel a combines all spiked virus recovery results and panel b separates recovery by virus. 

<br>

```{r volume, fig.height = 6}

clean_ww <- clean|>
  filter(water_type == "WW")

vol_ww <- clean_ww|>
  filter(!method %in% c("Amicon", "IP"))

glm_ww <- glm(rec ~ volume, data = vol_ww, family = Gamma(link = "log"))
lm_ww <- lm(log(rec) ~ as.factor(volume), data = vol_ww)

vol_comp_aov <- tibble(label = paste_anova_fun(anova = anova(lm_ww), log = TRUE),
                       row = 1)

pairwise.vol.ttest <- pairwise.t.test(log(vol_ww$rec), as.factor(vol_ww$volume))[["p.value"]]|>
  as_tibble(rownames = "volx")|>
  pivot_longer(2:3, names_to = "voly", values_to = "p.v")

vol_ww|>
  mutate(volume = fct_reorder(factor(volume), volume),
         row = row_number())|>
  left_join(vol_comp_aov, by = "row")|>
  ggplot(aes(volume, rec))+
  geom_boxplot(outlier.alpha = 0)+
  geom_beeswarm(alpha = 0.5)+
  geom_text(aes(x = 2, y = max(rec) + (0.07 * (max(rec) - min(rec))),
                label = label), size = 3)+
  labs(x = "Volume (ml)",
       y = "Recovery (%)") +
  
  vol_ww|>
  mutate(volume = fct_reorder(factor(volume), volume))|>
  ggplot(aes(volume, rec))+
  geom_boxplot(outlier.alpha = 0)+
  geom_beeswarm(alpha = 0.2, cex = 3)+
  facet_wrap(~virus, scales = "free")+
  labs(x = "Volume (ml)",
       y = "Recovery (%)")+
  theme(axis.title = element_blank(),
        axis.text.x = element_text(angle = 90)) +
  
  plot_layout(widths = c(1.3, 0.7)) + plot_annotation(tag_levels = "a")

```


Figure 3: Reduced starting volumes of wastewater improve viral recovery. Comparisons were made with an ANOVA after log transformation of recovery, followed by pairwise t tests; Pairwise comparisons found significant differences between all volumes with the holm adjustment method (p-value < 0.05: `r filter(pairwise.vol.ttest, p.v < 0.05)|>unite(vol_comp, c(volx, voly), sep = " ~ ")|>pull(vol_comp)|>paste0(collapse = "; ")`). Panel a combines recovery results and panel b separates recovery by virus. 


<br>

```{r, method}

clean_ww_15 <- clean_ww|>
  filter(volume == 15)

lm_meth <- lm(log(rec) ~ method, data = clean_ww_15)
meth_comp_aov <- tibble(label = paste_anova_fun(anova = anova(lm_meth), log = TRUE),
                       row = 1)


clean_ww_15|>
  mutate(method = fct_reorder(method, rec),
         row = row_number())|>
  left_join(meth_comp_aov, by = "row")|>
  ggplot(aes(method, rec))+
  geom_boxplot(outlier.alpha = 0)+
  geom_beeswarm(alpha = 0.5)+
  geom_text(aes(x = 3, y = max(rec) + (0.07 * (max(rec) - min(rec))),
                label = label), size = 3)+
  labs(x = "Concentration method",
       y = "Recovery (%)") +
  
  clean_ww_15|>
  mutate(method = fct_reorder(str_replace_all(method, "Amicon", "Ami"), rec))|>
  ggplot(aes(method, rec))+
  geom_boxplot(outlier.alpha = 0)+
  geom_beeswarm(alpha = 0.2)+
  facet_wrap(~virus, scale = "free")+
  labs(x = "Concentration method",
       y = "Recovery (%)") +
  
  plot_layout(heights = c(1.3, 0.7), nrow = 2, ncol = 1) + plot_annotation(tag_levels = "a")
  
```


Figure 4: Concentration methods influence virus recovery at a starting volume of 15ml. Panel a combines all recovery results with starting volumes of 15ml and panel b separates recovery by virus. Comparisons were made with an ANOVA after log transformation of recovery, followed by pairwise t tests; Pairwise comparisons found significant differences between 
